Forecasting the 2024 U.S. Presidential Election: Kamala Harris' Projected Victory

### 2024-11-04 paper/paper.qmd
committing text changes to paper.qmd

Data cleaning, analysis, modeling, and visualization for this paper were conducted using the statistical programming language `R` [@citeR] and the following packages: `tidyverse` package of @Tidyverse , `janitor` package of @Janitor for data cleaning, `kableExtra` package of @kableExtra for dynamic report generation and table formatting, `modelsummary` package of @Modelsummary for summarizing models, `usmap` package of @usmap for mapping U.S. states, `arrow` package of @arrow for data integration with Apache Arrow, `lubridate` package of @Lubridate for date handling, `rstanarm` package of @rstanarm and `brms` package of @brms for Bayesian regression modeling, `Matrix` package of @Matrix for handling sparse and dense matrices, `splines` package of @splines for regression splines, `caret`package of @caret for machine learning support, and `broom.mixed` package of @broommixed for tidying mixed models.

@@

@@

- **'State'**: Indicates the state where the polling took place, critical for understanding regional support dynamics.
- **'End Date'**: The date when the poll ended, which helps in analyzing trends over time.
- **'Start Date'**: The date when the poll started, to filter for Kamala Harris's late entry into the presidential race 
- **'Pollster'**: The organization conducting the poll, providing insight into the reliability and methodology of the polling data.
- **'Numeric Grade'**: A score assigned to each poll based on its methodology and historical accuracy, which assists in filtering out less reliable polls.
- **'Transparency Score'**: A measure indicating how transparent the pollster is regarding their methodology, further enhancing data credibility.
- **'Candidate Name'**: Identifies the running candidate, specifically focusing on Donald Trump and Kamala Harris.
- **Percentage of Support (pct)**: Represents the percentage of respondents supporting each candidate, which is the primary outcome variable of interest.

@@

```
@tbl-outcomevar reveals a total of 107 polls conducted, calculated as the total count of rows in our cleaned dataset. ‘Total polls’ reflects the total number of unique poll entries after filtering for only battleground states, polls with sufficient data quality, and polls that report support percentages for both Trump and Harris. The 107 polls conducted display an average support level of 47.801% for Trump and 47.999% for Harris. This suggests that based on the dataset, Trump has a leading advantage in winning the presidential election race by having a slighty higher average level of support than Harris. The highest percentage of support for Trump in the 107 polls was 51.700% while the lowest recorded support was 43%. For Harris, the highest percentage of support also reached 51.200%, with a minimum of 43.600%. These statistics illustrate the competitive landscape of the presidential election race, without a clear leading candidate in the current race.

It is important to note that since **'pct'** stands for percentage, there may be a misconception that the values for Trump and Harris will always add up to 100%. For instance, in the summary statistics presented in Table 1, the average support for Donald Trump is 47.801%, while the average support for Kamala Harris is 47.999%. This results in a combined average support of 95.800%, indicating that there is a significant portion of respondents who are either undecided or supporting other candidates. This highlights the competitive nature of the election and emphasizes the necessity of examining not only the percentages attributed to the main candidates but also the context of voter preferences that could influence the final outcome.

@@

#### State
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-state-distribution
#| fig-cap: "Distribution of Observations for Each State in Polling Data. Illustrates the number of polling observations across key battleground statesn. Pennsylvania has the highest count of observations, suggesting a more robust polling presence, while Nevada shows fewer observations."

# Visualize the distribution of the state variable
ggplot(clean_poll_data, aes(x = state)) +
  geom_bar(fill = "steelblue") +  # Bar plot with specified fill color
  labs(
    x = "State",
    y = "Count of Observations"
  ) +
  theme_minimal() +  # Use a minimal theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for better readability
    plot.title = element_text(hjust = 0.5)  # Center the plot title
  )
```
The state (**'state'**) variable indicates the U.S. state where the poll was conducted or targeted. In the context of the U.S. presidential election, the outcome is frequently determined by several key battlegroud states. In the current 2024 election cycle, there are seven battleground states that are pivotal in the contest between Kamala Harris, representing the Democratic Party, and Donald Trump, representing the Republican Party. To enhance our model and refine our predictions, we focused exclusively on these seven battleground states. The polling results from each of these states play a crucial role in predicting electoral outcomes, as the percentage of support is evaluated on a state-by-state basis. The distribution of the state variable is visualized in @fig-state-distribution and indicates that some states, such as Pennsylvania, North Carolina and Wisconsin have more polling observations compared to others, like Nevada. This variability reflects differences in polling frequency and interest among these battleground states. States with more polling observations provide a larger sample size, potentially leading to more stable and reliable predictions for those states in the model.

#### End Date {#sec-end-date}
End date (**'end_date'**) represents the date the poll ended, essentially the date the poll stopped collecting data from electorates. In the context of our paper, the end date of each poll serves as one of the predictive variables and is used to assess changes in voter support over time. Including end date as a variable in our model can help us identify patterns in voter support as Election Day approaches, potentially reflecting the impact of campaign events, debates, or other political developments. @fig-end-date-distribution displays the distribution of polling end dates from July 2024 to October 2024.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-end-date-distribution
#| fig-cap: "Distribution of Polling End Dates. This histogram displays the frequency of polling end dates for key battleground states, highlighting peaks in polling activity around late September."

@@

```
We filtered for the earliest end date as July 2024 since Harris officially announced her presidential campaign on July 21, 2024. Therefore, any polls that has an end date before July would not contain polling results for Harris. The data from Project 538 [@FiveThirtyEight] is last updated on October 15th, 2024, thus the latest end date being October 14th, 2024. Furthermore, @fig-end-date-distribution also reveals that polling activity is not evenly distributed over time. There are clear peaks where the number of polls conducted is higher, notably in early September, late September, as well as early October. This concentration of polls likely corresponds to period of heightened political interest or significant campaign events that prompt increased polling.

@@

#### Numeric Grade
Numeric Grade (**'numeric_grade'**) represents the quality of each poll, as rated by Project 538 [@FiveThirtyEight], based on factors such as methodology, transparency, and historical accuracy. In the context of our paper, numeric grade allows us to weigh the reliability of different polls. By including numeric grade as a variable in our model, we aim to account for variations in poll quality, ensuring that polls with higher ratings (indicating higher reliability) contribute more significantly to our predictions of voter support. 
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-numgrade-distribution
#| fig-cap: "Density plot illustrating the distribution of numeric grades assigned to pollsters within the polling data. Highlights the frequency of numeric grades, indicating key concentrations that suggest variability in pollster quality. "

@@

```
@fig-numgrade-distribution displays the distribution of numeric grades for the polls included in our analysis, ranging from 2.5 to 3.0. Polls with higher grades are generally seen as more methodologically rigorous and transparent, which helps improve the reliability of our model’s predictions. The distribution shows several peaks, indicating that certain grade levels (around 2.7, 2.8, and 3.0) are more common among the polling data we collected. This variability in numeric grade highlights the diversity in polling quality within our dataset. The presence of peaks suggests that some pollsters consistently receive higher ratings, likely due to their use of more rigorous methods or transparent reporting standards. Conversely, the valleys in the distribution may indicate fewer polls from sources with lower numeric grades, reflecting a potential tendency for lower-quality polls to be less prevalent or less frequently conducted. By accounting for numeric grade, we can improve the accuracy of predictions by emphasizing data from more reliable sources. This consideration is crucial, as polling quality can impact the stability of the model’s forecasts, especially in closely contested states where polling accuracy is essential for reliable predictions.

#### Transparency Score
Transparency Score (**'transparency_score'**) represents the degree of methodological transparency provided by each pollster, rated on a scale from 0 to 10 by Project 538 [@FiveThirtyEight]. This score reflects the extent to which pollsters disclose details about their data collection and analysis methods, such as sample size, weighting, and margin of error. In the context of our paper, transparency score is an essential variable for evaluating the reliability of polling data, as higher transparency scores generally indicate a greater level of methodological rigor and openness. @fig-transparency-score-distribution displays the distribution of transparency scores across the polling data for key battleground states. The plot shows that transparency scores cluster heavily around ratings of  7 and 9, suggesting that a majority of polls in our dataset come from sources that maintain a relatively high level of transparency. There are smaller peaks at lower scores (5 and 10), indicating that while some pollsters offer limited transparency, a few polls also achieve perfect transparency scores.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-transparency-score-distribution
#| fig-cap: "Histogram displaying the distribution of transparency scores among pollsters in the polling data. Indicates the frequency of various transparency scores, highlighting a concentration of scores around 7 and 9, which suggests a predominance of high transparency among the assessed pollsters. "

@@

```
This distribution is important for understanding the credibility of our data sources. Polls with higher transparency scores are more likely to provide consistent and reliable results, as they openly disclose their methodologies, allowing us to assess potential biases or limitations in their data. In contrast, polls with lower transparency scores may lack sufficient methodological detail, making it challenging to gauge their accuracy. Incorporating transparency scores into our model helps differentiate between high- and low-transparency polls. This adjustment is crucial as transparency can influence the stability of the model’s predictions, especially in closely contested states where the reliability of polling data may impact the accuracy of our forecasted election outcomes.

@@

## Model Setup

@@

## Model Justification
The Bayesian Beta Regression model with a logit link function is the most suitable model for predicting voter support for Kamala Harris and Donald Trump in the 2024 U.S. Presidential Election. Our primary outcome variable is the percentage of support for each candidate in seven key battleground states, which is naturally bounded between 0% and 100%. The Beta distribution is ideal for modeling this proportion data, as it allows us to avoid unrealistic predictions outside the (0, 1) range. To capture non-linear trends in voter support as Election Day approaches, we employe natural splines with **'end_date_num'**, which represents the number of days from earliest **'end_date'** to latest **'end_date'**. Additionally, incorporating state-specific random effects acknowledges the variations in voter preferences across different regions, allowing for more accurate modeling of baseline support levels. The Bayesian framework enhances our model by enabling the incorporation of prior information, leading to more stable estimates, especially in the context of sparse polling data. Furthermore, the model's structure allows for clear interpretation in terms of victory probabilities, aligning well with our goals of forecasting electoral outcomes. Further details are discussed in [Appendix -@sec-model-details].

## Model Summary
### Trump Model Summary
@tbl-trump-model-summary displays fixed effects estimates for several key terms that capture the influence of time on his support in the battleground states. The intercept, with an estimate of -0.0923, represents the baseline log-odds of Trump’s support when other predictors are held constant. This negative intercept suggests a slightly lower baseline level of support. The term nsenddatenumdfEQ42, with an estimate of 0.0803 and a statistically significant positive confidence interval (0.0179 to 0.1426), indicates an increase in support around this time period. Similarly, nsenddatenumdfEQ44 has an estimate of 0.0789 and a confidence interval above zero (0.0014 to 0.1589), also showing a positive effect on support. Other terms such as nsenddatenumdfEQ41 and nsenddatenumdfEQ43 have less significant impacts, suggesting that certain time periods were more influential than others in shaping Trump’s support. The standard errors for the fixed effects terms range from 0.0306 to 0.0676 for nsenddatenumdfEQ43. The relatively low standard error of 0.0306 on the intercept suggests that the baseline estimate is quite stable, meaning we can be more confident about this baseline support level. However, the standard error for nsenddatenumdfEQ43 (0.0676) is higher than for other terms, indicating that this estimate is less precise.
```{r}
#| echo: false
#| eval: true
#| label: tbl-trump-model-summary
#| tbl-cap: "Coefficient estimates from the Bayesian Beta regression model for Donald Trump, showing the impact of the natural spline transformation of end date on voter support. The intercept indicates baseline log-odds, while spline terms reflect nonlinear trends in voter sentiment, with positive coefficients suggesting increased support as Election Day approaches."
#| warning: false

@@

# Display the Trump model summary table
knitr::kable(trump_summary) %>%
  kable_styling(full_width = FALSE, position = "center")
```
These results imply that support for Trump varied over time in ways that may correspond to specific campaign events or other influences affecting voter sentiment. The significance of terms like nsenddatenumdfEQ42 and nsenddatenumdfEQ44 suggests that changes in voter preference could be captured effectively by this model. And by checking standard errors, helping us determine how much we can rely on those data.

### Harris Model Summary
@tbl-harris-model-summary similarly shows the impact of time on her support in the battleground states. The intercept for the Harris model is estimated at -0.1508, suggesting a slightly lower baseline support level compared to Trump(-0.0923). The term nsenddatenumdfEQ41 has an estimate of 0.1225 and a confidence interval 0.0553 to 0.1903, indicating a significant increase in support during this period. Another positive and significant effect is seen in nsenddatenumdfEQ43, with an estimate of 0.1191 and confidence interval from -0.0071 to 0.2417, which suggests a trend toward increased support. While other terms stay positive, they have less impacts on support. Additionally, the standard errors for fixed effects are generally smaller, ranging from 0.0279 for the intercept to 0.0631 for nsenddatenumdfEQ43. The larger standard error for nsenddatenumdfEQ43 (0.0631) suggests that there would be some uncertainty in how this time point influences support, though the positive estimate still points toward an upward trend. In comparison to the Trump model, the standard errors in the Harris model are slightly smaller for most terms, indicating potentially more consistent support trends over time in the data analyzed.
```{r}
#| echo: false
#| eval: true
#| label: tbl-harris-model-summary
#| tbl-cap: "Coefficient estimates from the Bayesian Beta regression model for Kamala Harris, illustrating the effects of the natural spline transformation of end date on voter support. The intercept represents the baseline log-odds, while the spline terms indicate nonlinear trends in voter sentiment, with positive coefficients suggesting an increase in support as Election Day nears."
#| warning: false

@@

```
These results highlight that Harris's support also fluctuated over time, with nsenddatenumdfEQ41 and nsenddatenumdfEQ43 marking periods of notable increases. Helping us understand the pattern, this model predicts Harris's probability of victory and provides insights for future changes in support as the election nears.

# Results {#sec-results}
In this section, we present the results of our analysis, focusing on the electoral support for Donald Trump and Kamala Harris across key battleground states during the 2024 Presidential Election cycle. We begin by examining the state-level polling averages for both candidates, which provide a snapshot of voter preferences as captured by various pollsters. Next, we explore how polling averages have evolved over time, highlighting trends that may influence election outcomes. We then assess the predicted probabilities of Trump winning on Election Day, followed by a detailed look at the likelihood of winning by state. Finally, we visualize the predicted winner by state on a US map, summarizing our findings and their implications for the electoral landscape.

@@

Social desirability bias and non-response bias add further uncertainty, especially regarding Trump’s support base. Social stigma may deter some voters from expressing conservative support openly, leading to under-reported Trump support in polls. Additionally, Trump’s supporters often have lower social trust, making them less likely to participate in polls altogether [@discussion-article3]. If certain groups systematically avoid participation, polling data may under-represent these voters, resulting in an underestimation of Trump’s support in our model.

On the other hand, Harris faces potential challenges such as the Bradley effect and gender-related bias. As a Black female candidate, Harris may encounter undecided voters leaning against her due to underlying biases, similar to what Hillary Clinton faced in 2016 [@discussion-article3]. If undecided voters ultimately vote against Harris, our model may overestimate her support based on polling data alone, affecting prediction accuracy.

@@

# Additional Data Details {#sec-data-details}
## Data Cleaning {#sec-data-cleaning}
The data cleaning process for our analysis focused on refining the raw polling data obtained from Project 538's polling project [@FiveThirtyEight].  The raw data contains polling information for various U.S. states, but for our specific analysis, we concentrated on seven designated battleground states: Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin. This is due to the fact that these battleground states have been recognized as pivotal in determining this year's election outcome.

To prepare the data for modeling, we performed several key steps. First, we filtered the data to retain only relevant columns, including state, polling dates, pollster information, numeric grades, transparency scores, candidate names, and their respective support percentages. This filtering process ensured that we excluded any unnecessary data that could complicate our analysis. Next, we implemented thresholds for the numeric grade (minimum of 2.5) and transparency score (greater than 5) to ensure that only high-quality polling data was considered. We also converted polling dates into a standardized date format and created a new numeric variable representing the number of days since the earliest polling date. Additionally, we handled specific data inconsistencies, such as replacing "Nebraska CD-2" with "Nebraska." The polling percentages were then aggregated by averaging the values for each candidate while preserving other important attributes such as pollster and state. This aggregation allowed us to focus on polling data that represented a balanced view of voter support across the selected battleground states. The cleaned dataset was then saved in a parquet format for efficient processing in subsequent analyses.

To ensure the integrity of our cleaned polling dataset, we implemented a comprehensive data testing process. This included verifying the presence of all required columns, checking data types for consistency, and ensuring there were no missing values in critical fields. We validated that candidate support percentages fell within the logical range of 0 to 100 and confirmed that numeric grades and transparency scores met predefined thresholds. By rigorously applying these tests, we ensured that the dataset is reliable and suitable for subsequent modeling and analysis, providing a solid foundation for our exploration of voter behavior in the battleground states.

@@

### Model Diagnostics - Harris Model
The posterior predictive check for the Harris model illustrates the relationship between the observed support percentages $y$ and the predicted support values $y_{\text{rep}}$. @fig-harris-check displays a density estimate for both the actual observed data (depicted in dark blue) and the predicted values generated by the model (shown in light blue). The purpose of this diagnostic check is to evaluate how well the model captures the distribution of observed support for Kamala Harris.
```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-harris-check
#| fig-cap: "Posterior predictive check for Kamala Harris's model, illustrating the distribution of observed support percentages $y$ against the predicted support values $y_{\text{rep}}$. The graph shows a strong correspondence between the observed and predicted distributions, indicating that the model accurately captures the dynamics of voter support for Harris in the given dataset."

>>>>

@@

Data cleaning and analysis are conducted by employing the following packages: `tidyverse` package of @Tidyverse, `janitor` package of @Janitor, `lubridate` package of @Lubridate,  `arrow` package of @arrow

@@

- **'State'**: Indicates the state where the polling took place, critical for understanding regional support dynamics.
- **'End Date'**: The date when the poll ended, which helps in analyzing trends over time.
- **'Start Date'**: The date when the poll started, to filter for Kamala Harris's late entry into the presidential race 
- **'Pollster'**: The organization conducting the poll, providing insight into the reliability and methodology of the polling data.
- **'Numeric Grade'**: A score assigned to each poll based on its methodology and historical accuracy, which assists in filtering out less reliable polls.
- **'Transparency Score'**: A measure indicating how transparent the pollster is regarding their methodology, further enhancing data credibility.
- **'Candidate Name'**: Identifies the running candidate, specifically focusing on Donald Trump and Kamala Harris.
- **Percentage of Support ('pct')**: Represents the percentage of respondents supporting each candidate, which is the primary outcome variable of interest.

@@

```
@tbl-outcomevar reveals a total of 107 polls conducted, representing unique poll entries filtered for battleground states, sufficient data quality, and reported support percentages for both Trump and Harris. Among these, the average support level is 47.801% for Trump and 47.999% for Harris, indicating a slight advantage for Trump in the presidential race. The highest recorded support for Trump is 51.700%, while the lowest is 43%; for Harris, the highest is 51.200% and the lowest is 43.600%. These figures illustrate a competitive landscape without a clear leading candidate. It is essential to note that the percentages for Trump and Harris do not sum to 100%, as the combined average of 95.800% highlights a significant portion of respondents who are undecided or supporting other candidates. This highlights the competitive nature of the election and emphasizes the necessity of examining not only the percentages attributed to the main candidates but also the context of voter preferences that could influence the final outcome.

@@

#### State
The state (**'state'**) variable indicates the U.S. state where the poll was conducted or targeted. In the context of the U.S. presidential election, the outcome is frequently determined by several key battlegroud states. In the current 2024 election cycle, there are seven battleground states that are pivotal in the contest between Kamala Harris, representing the Democratic Party, and Donald Trump, representing the Republican Party. To enhance our model and refine our predictions, we focused exclusively on these seven battleground states. The polling results from each of these states play a crucial role in predicting electoral outcomes, as the percentage of support is evaluated on a state-by-state basis.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-state-distribution
#| fig-cap: "Distribution of Observations for Each State in Polling Data. Illustrates the number of polling observations across key battleground statesn. Pennsylvania has the highest count of observations, suggesting a more robust polling presence, while Nevada shows fewer observations."

# Visualize the distribution of the state variable
ggplot(clean_poll_data, aes(x = state)) +
  geom_bar(fill = "steelblue") +  # Bar plot with specified fill color
  labs(
    x = "State",
    y = "Count of Observations"
  ) +
  theme_minimal() +  # Use a minimal theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for better readability
    plot.title = element_text(hjust = 0.5)  # Center the plot title
  )
```
The distribution of the state variable is visualized in @fig-state-distribution and indicates that some states, such as Pennsylvania, North Carolina and Wisconsin have more polling observations compared to others, like Nevada. This variability reflects differences in polling frequency and interest among these battleground states. States with more polling observations provide a larger sample size, potentially leading to more stable and reliable predictions for those states in the model.

#### End Date {#sec-end-date}
End date (**'end_date'**) marks the conclusion of data collection for each poll, serving as a predictive variable in our analysis of changes in voter support over time. This inclusion helps identify patterns as Election Day approaches, potentially reflecting the impact of campaign events, debates, or political developments.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-end-date-distribution
#| fig-cap: "Distribution of Polling End Dates. This histogram displays the frequency of polling end dates for key battleground states, highlighting peaks in polling activity around late September."

@@

```
@fig-end-date-distribution shows polling end dates from July 2024 to October 2024, with the earliest date set to July 2024, coinciding with Kamala Harris's official presidential campaign announcement on July 21, 2024. The latest end date, October 14, 2024, reflects the last data update from Project 538 [@FiveThirtyEight] on October 15, 2024. The distribution reveals that polling activity is not evenly spread, with notable peaks in early September, late September, and early October, likely corresponding to periods of heightened political interest or significant campaign events that drive increased polling activity.

@@

#### Numeric Grade
Numeric Grade (**'numeric_grade'**) assesses the quality of each poll, as rated by Project 538 [@FiveThirtyEight], based on factors like methodology, transparency, and historical accuracy. In our paper, this variable allows us to weigh the reliability of different polls, ensuring that those with higher ratings contribute more significantly to our voter support predictions. 
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-numgrade-distribution
#| fig-cap: "Density plot illustrating the distribution of numeric grades assigned to pollsters within the polling data. Highlights the frequency of numeric grades, indicating key concentrations that suggest variability in pollster quality. "

@@

```
@fig-numgrade-distribution illustrates the distribution of numeric grades in our analysis, which ranges from 2.5 to 3.0. Higher grades indicate more methodologically rigorous and transparent polls, enhancing the reliability of our model's predictions. The distribution shows several peaks around grades 2.7, 2.8, and 3.0, suggesting that certain pollsters consistently achieve higher ratings due to better methodologies or transparency. Conversely, valleys in the distribution may reflect a scarcity of polls from lower-rated sources. By incorporating numeric grade, we enhance prediction accuracy by prioritizing data from more reliable sources, which is particularly crucial in closely contested states where polling accuracy is essential for reliable forecasts.

#### Transparency Score
Transparency Score (**'transparency_score'**) quantifies the methodological transparency of each pollster on a scale from 0 to 10, as rated by Project 538 [@FiveThirtyEight]. This score reflects the extent of disclosure regarding data collection and analysis methods, including sample size, weighting, and margin of error. In our paper, the transparency score is crucial for assessing the reliability of polling data; higher scores typically indicate greater methodological rigor. 
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-transparency-score-distribution
#| fig-cap: "Histogram displaying the distribution of transparency scores among pollsters in the polling data. Indicates the frequency of various transparency scores, highlighting a concentration of scores around 7 and 9, which suggests a predominance of high transparency among the assessed pollsters. "

@@

```
@fig-transparency-score-distribution illustrates the distribution of transparency scores in our dataset for key battleground states, showing a concentration around scores of 7 and 9, indicating that most polls come from sources with relatively high transparency. Smaller peaks at lower scores (5) and a perfect score (10) suggest that while some pollsters provide limited transparency, a few achieve full disclosure. This distribution is vital for evaluating data credibility, as polls with higher transparency scores are more likely to yield consistent and reliable results, enabling us to identify potential biases. Incorporating transparency scores into our model helps differentiate between high- and low-transparency polls, which is essential for ensuring the stability of our predictions, particularly in closely contested states where polling accuracy is critical.

@@

## Model Setup
Modeling processes are conducted by employing the following packages: `rstanarm` package of @rstanarm, `brms` package of  @brms for Bayesian regression modeling, `Matrix` package of @Matrix for handling sparse and dense matrices, `splines` package of @splines for regression splines, `caret` package of @caret for machine learning support, and `broom.mixed` package of @broom.mixed for tidying mixed models.

@@

## Model Justification
The Bayesian Beta Regression model with a logit link function is the most suitable model for predicting voter support for Kamala Harris and Donald Trump in the 2024 U.S. Presidential Election. Our primary outcome variable is the percentage of support for each candidate in seven key battleground states, which is naturally bounded between 0% and 100%. The Beta distribution is ideal for modeling this proportion data, as it allows us to avoid unrealistic predictions outside the (0, 1) range. To capture non-linear trends in voter support as Election Day approaches, we employe natural splines with **'end_date_num'**, which represents the number of days from earliest **'end_date'** to latest **'end_date'**. Additionally, incorporating state-specific random effects acknowledges the variations in voter preferences across different regions, allowing for more accurate modeling of baseline support levels. The Bayesian framework enhances our model by enabling the incorporation of prior information, leading to more stable estimates, especially in the context of sparse polling data. Furthermore, the model's structure allows for clear interpretation in terms of victory probabilities, aligning well with our goals of forecasting electoral outcomes. Details are further discussed in [Appendix -@sec-model-details].

## Model Summary
### Trump Model Summary
The summary in @tbl-trump-model-summary presents fixed effects estimates reflecting the influence of time on Trump's support in battleground states. The intercept, estimated at -0.0923, indicates a slightly lower baseline log-odds of support when other predictors are held constant. The term nsenddatenumdfEQ42 shows an estimate of 0.0803 with a statistically significant confidence interval (0.0179 to 0.1426), suggesting an increase in support during this period. Similarly, nsenddatenumdfEQ44 has an estimate of 0.0789 and a confidence interval above zero (0.0014 to 0.1589), also indicating a positive effect. In contrast, nsenddatenumdfEQ41 and nsenddatenumdfEQ43 exhibit less significant impacts, suggesting variability in time periods' influence on support. Standard errors for the fixed effects range from 0.0306 to 0.0676, with the lower standard error of 0.0306 for the intercept indicating a stable baseline estimate. However, the higher standard error of 0.0676 for nsenddatenumdfEQ43 reflects less precision in that estimate. Overall, these results imply that Trump's support fluctuated over time, likely in response to specific campaign events, with significant terms indicating changes in voter preference effectively captured by the model. Standard errors further inform the reliability of these estimates.
```{r}
#| echo: false
#| eval: true
#| label: tbl-trump-model-summary
#| tbl-cap: "Coefficient estimates from the Bayesian Beta regression model for Donald Trump, showing the impact of the natural spline transformation of end date on voter support. The intercept indicates baseline log-odds, while spline terms reflect nonlinear trends in voter sentiment, with positive coefficients suggesting increased support as Election Day approaches."
#| warning: false

@@

# Display the Trump model summary table
knitr::kable(trump_summary) %>%
  kable_styling(full_width = FALSE, position = "center")
```
Overall, these results imply that Trump's support fluctuated over time, likely in response to specific campaign events, with significant terms indicating changes in voter preference effectively captured by the model. Standard errors further inform the reliability of these estimates.

### Harris Model Summary
@tbl-harris-model-summary similarly shows the impact of time on her support in the battleground states. The intercept is estimated at -0.1508, indicating a lower baseline support level compared to Trump's intercept of -0.0923. The term nsenddatenumdfEQ41 shows an estimate of 0.1225 with a confidence interval of 0.0553 to 0.1903, reflecting a significant increase in support during this period. Another notable effect is observed in nsenddatenumdfEQ43, which has an estimate of 0.1191 and a confidence interval from -0.0071 to 0.2417, suggesting a trend toward increased support. While other terms remain positive, they have less impact. The standard errors for fixed effects range from 0.0279 for the intercept to 0.0631 for nsenddatenumdfEQ43, with the larger standard error for nsenddatenumdfEQ43 indicating some uncertainty about its influence on support, despite the positive estimate suggesting an upward trend. Compared to the Trump model, the Harris model's standard errors are generally smaller, indicating more consistent support trends over time.
```{r}
#| echo: false
#| eval: true
#| label: tbl-harris-model-summary
#| tbl-cap: "Coefficient estimates from the Bayesian Beta regression model for Kamala Harris, illustrating the effects of the natural spline transformation of end date on voter support. The intercept represents the baseline log-odds, while the spline terms indicate nonlinear trends in voter sentiment, with positive coefficients suggesting an increase in support as Election Day nears."
#| warning: false

@@

```
These results reveal fluctuations in Harris's support, particularly during the significant periods marked by nsenddatenumdfEQ41 and nsenddatenumdfEQ43, which aid in predicting her probability of victory and potential shifts in support as the election approaches.

# Results {#sec-results}
In this section, we present the analysis of electoral support for Donald Trump and Kamala Harris in key battleground states during the 2024 Presidential Election cycle. We start by examining state-level polling averages for both candidates, providing insights into voter preferences as reported by various pollsters. We then analyze the evolution of these polling averages over time, identifying trends that could affect election outcomes. Following this, we assess the predicted probabilities of Trump winning on Election Day and examine the likelihood of winning by state. Finally, we visualize the predicted winner by state on a US map, summarizing our findings and their implications for the electoral landscape. Visualizations of results are generated by employing the following packages: `usmap` package of @usmap for mapping U.S. states, `kableExtra` package of @kableExtra for dynamic report generation and table formatting, `modelsummary` package of @Modelsummary for summarizing models.

@@

Social desirability bias and non-response bias add further uncertainty, especially regarding Trump’s support base. Social stigma may deter some voters from expressing conservative support openly, leading to under-reported Trump support in polls. Additionally, Trump’s supporters often have lower social trust, making them less likely to participate in polls altogether [@discussion-article3]. If certain groups systematically avoid participation, polling data may under-represent these voters, resulting in an underestimation of Trump’s support in our model. On the other hand, Harris faces potential challenges such as the Bradley effect and gender-related bias. As a Black female candidate, Harris may encounter undecided voters leaning against her due to underlying biases, similar to what Hillary Clinton faced in 2016 [@discussion-article3]. If undecided voters ultimately vote against Harris, our model may overestimate her support based on polling data alone, affecting prediction accuracy.

@@

# Additional Data Details {#sec-data-details}
## Data Cleaning {#sec-data-cleaning}
The data cleaning process for our analysis focused on refining the raw polling data obtained from Project 538's polling project [@FiveThirtyEight]. We concentrated on seven key battleground states—Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin—recognized as pivotal for this year's election outcome. Key steps included filtering relevant columns such as state, polling dates, pollster information, numeric grades, transparency scores, candidate names, and support percentages to eliminate unnecessary data. We set thresholds for numeric grades (minimum of 2.5) and transparency scores (greater than 5) to ensure the inclusion of high-quality polling data. Polling dates were standardized, and we created a numeric variable representing the number of days since the earliest polling date. Additionally, we addressed data inconsistencies, like replacing "Nebraska CD-2" with "Nebraska," and aggregated polling percentages by averaging candidate support while preserving essential attributes. The cleaned dataset was saved in parquet format for efficient processing in subsequent analyses.

To ensure the integrity of our cleaned data, we implemented a comprehensive data testing process. This involved verifying the presence of required columns, checking data types for consistency, and ensuring no missing values in critical fields. We validated that candidate support percentages fell within the logical range of 0 to 100 and confirmed that numeric grades and transparency scores met our predefined thresholds. By rigorously applying these tests, we established that the data is reliable and suitable for modeling and analysis, providing a solid foundation for exploring voter behavior in the selected battleground states.

@@

### Posterior Predictive Check for the Harris Model
The posterior predictive check for the Harris model illustrates the relationship between the observed support percentages $y$ and the predicted support values $y_{\text{rep}}$. @fig-harris-check displays a density estimate for both the actual observed data (depicted in dark blue) and the predicted values generated by the model (shown in light blue). The purpose of this diagnostic check is to evaluate how well the model captures the distribution of observed support for Kamala Harris.
```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-harris-check
#| fig-cap: "Posterior predictive check for Kamala Harris's model, illustrating the distribution of observed support percentages $y$ against the predicted support values $y_{\text{rep}}$. The graph shows a strong correspondence between the observed and predicted distributions, indicating that the model accurately captures the dynamics of voter support for Harris in the given dataset."
